\documentclass{article}

\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{tikz}
\usepackage[plain]{algorithm}
\usepackage{algpseudocode}

\usetikzlibrary{automata,positioning}

%
% Basic Document Settings
%

\topmargin=-0.45in
\evensidemargin=0in
\oddsidemargin=0in
\textwidth=6.5in
\textheight=9.0in
\headsep=0.25in

\linespread{1.1}

\pagestyle{fancy}
\lhead{\hmwkAuthorName}
\chead{} % empty
\rhead{\hmwkHeaderTitle}
\cfoot{\thepage}

\renewcommand\headrulewidth{0.4pt}
\renewcommand\footrulewidth{0pt}

\setlength\parindent{0pt}

%
% Homework Details
%

\newcommand{\hmwkTitle}{Homework\ 4}
\newcommand{\hmwkDueDate}{Tuesday, February 24, 2026,\ 11{:}59 pm (Thai time)}
\newcommand{\hmwkClass}{Data Structure\ and Abstractions}
\newcommand{\hmwkClassTime}{Section 2}
\newcommand{\hmwkClassInstructor}{Kanat Tangwongsan}
\newcommand{\hmwkAuthorName}{\textbf{Jiraroj Wiruchpongsanon (6781617)}}

% What appears in the top-right header
\newcommand{\hmwkHeaderTitle}{Data Structure and Abstractions: Homework 4}

%
% Title Page
%

\title{
    \vspace{2in}
    \textmd{\textbf{\hmwkClass:\ \hmwkTitle}}\\
    \normalsize\vspace{0.1in}\small{Due\ on\ \hmwkDueDate}\\
    \vspace{0.1in}\large{\textit{\hmwkClassInstructor\ \hmwkClassTime}}
    \vspace{3in}
}

\author{\hmwkAuthorName}
\date{}

\renewcommand{\part}[1]{\textbf{\large Part \Alph{partCounter}}\stepcounter{partCounter}\\}

%
% Helper Commands
%

\newcounter{partCounter}
\newcommand{\solution}{\textbf{\large Solution}}

\newtheorem*{theorem}{Theorem}

\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\begin{document}

\maketitle
\newpage

\section*{Task 1: Missing Tile (6 points)}

For this task, save your work in \texttt{smallman/MissingTile.java}. There are four ways an L-shaped triomino can be arranged, labeled by the missing corner: NE, SE, SW, or NW.

\begin{theorem}
Any $2^n \times 2^n$ grid with one painted cell can be tiled using L-shaped triominoes such that the entire grid is covered by triominoes but no triomino overlaps either another triomino or the painted cell.
\end{theorem}

\textbf{Subtask I:} Re-prove the theorem above, in your own words, by induction on $n$ so that you fully understand how it works.

\medskip
\solution

\begin{proof}

    we claim that

    \[
        P(n) \equiv \forall n \ge 1 \text{ any $2^n \times 2^n$ grid with one painted cell can be fully tiled using L-shaped triminoes in any orientation}
    \]

    for the \textbf{base case} where $n_0 = 1$

    \begin{align*}
        P(1) &\equiv \text{a $2^1 \times 2^1$ grid with one painted entry could be covered by a single triminos} \\
        &\text{\hspace{1em} (as one of the four possible arrangement would work)}
    \end{align*}

    then, we assume P(k) for arbitrary $(k \ge n_0)$, we prove $P(k+1)$

    \vspace{1em}

    we split $2^{k+1} \tiems 2^{k+1}$ into 4 quadrant: $Q_1, Q_2, Q_3 \text{ and } Q_4$, as we could positioned the (single) painted entry by it's quadrant --- which it has to be in one of the four quadrant.

    \vspace{1em}

    this quadrant that contains the painted entry would be the size of $\frac{2^{k+1}}{2} \times \frac{2^{k+1}}{2} = 2^k \times 2^k$, which is just $P(k)$ --- that from our inductive hypothesis, holds true.

    \vspace{1em}

    now, we focus on the remaining three quadrant, these three, if we view each of them as a single cell, will be arranged in one of the four possible L-shaped triminoes arrangement. if we put in a single L-shaped trimino on the apex of those three $2^k \times 2^k$ grid, each of those three grid would have one used (painted) entry --- which is just the set-up of $P(k)$ that we assume to be true already; and therefore, by inductive hypothesis, these three grid, together, will be tilable with L-shaped triminoes.

    \vspace{1em}
    and since P(k+1) follows P(k), the \textbf{induction case} is satisfied, thus we can conclude the given predicate is true

\end{proof}

\newpage 

\section*{Task 4: Tail Sum of Squares (2 points)}

Consider the following snippet of Java code:
\begin{verbatim}
int sumHelper(int n , int a) {
    if (n==0) return a;
    else return sumHelper(n-1, a + n*n);
}
int sumSqr(int n) { return sumHelper(n, 0); }
\end{verbatim}

Your Task: Prove that for $n \ge 1$, \texttt{sumSqr(n)} returns $1^2 + 2^2 + 3^2 + \cdots + n^2$. To prove this, use induction to show that \texttt{sumHelper} computes the right thing. (Hint: Think about how we proved \texttt{fact\_helper} in class.)

\medskip
\solution

\begin{proof}
as sumSqr(int n) only calls sumHelper(n, 0) for $n \ge 1$, we focus on sumHelper(n, 0), let the claim

\[
P(n) \equiv \forall a,\; \text{ sumHelper(n, a) } = a + \sum_{i=1}^n i^2
\]

now, we prove this using induction; for the \textbf{base case} where $n_0 = 0$

\begin{align*}
    P(0) \equiv \text{ sumHelper(0, a)} 
    &= a \\
    &= a + \sum_{i=1}^{0} i^2
\end{align*}

then, we assume P(k) for arbitrary $(k \ge n_0)$, we prove $P(k+1)$

\begin{align*}
    P(k+1) &\equiv \text{ sumHelper(k+1, a)} \\
    &= \text{ sumHelper(k, a + $(k+1)^2)$} \\
    &= (a + (k+1)^2) + \sum_{i=1}^{k} i^2 \\
    &= a + \sum_{i=1}^{k} i^2 + (k+1)^2 \\
    &= a + \sum_{i=1}^{k+1} i^2
\end{align*}

since P(k+1) follows P(k) for arbitrary $a$, the \textbf{induction case} is satisfied, thus we can conclude the given predicate is true.

therefore, 
\[
\text{ sumSqr(n)} = \text{ sumHelper(n, 0)} = \sum_{i=1}^{n} i^2
\]

\end{proof}

\section*{Task 5: Mysterious Function (2 points)}

Consider the following Python function \texttt{foo}, which takes as input an integer $n \ge 1$ and returns a tuple of length 2 of integers:
\begin{verbatim}
def foo(n):
    assert n>=1
    if n == 1:
        return (1, 2)
    else:
        p, q = foo(n-1)
        return (q + p*n*(n+1), q*n*(n+1))
\end{verbatim}

Prove that for $n \ge 1$, \texttt{foo(n)} returns a pair $(p, q)$ such that
\[
    \frac{p}{q} = 1 - \frac{1}{n + 1}.
\]
(Hint: Use induction on $n$.)

\medskip
\solution

\begin{proof}

    we claim that

    \[
        P(n) \equiv \forall n \ge 1, \text{ foo(n) returns pq such that } \frac{p}{q} = 1 - \frac{1}{n+1}
    \]

    for the \textbf{base case} where $n_0 = 1$

    \begin{align*}
        P(1) &\equiv \text{ foo(1) would returns } (1, 2) \simeq (p, q) \\
        &\mapsto \frac{p}{q} = \frac{1}{2} = 1 - \frac{1}{1+1}
    \end{align*}

    then, we assume P(k) for arbitrary $(k \ge n_0)$, we prove $P(k+1)$

    \begin{align*}
        P(k+1) &\equiv \text{ foo(k+1)} \text{; $p_0, q_0$ will be the result of foo(k)}
    \end{align*}

    therefore 

    \[
    \frac{p_0}{q_0} = \frac{k}{k+1} \quad \text{ as induction hypothesis}
    \]

    foo(k+1) would return

    \[
    (q_0 + p_0(k+1)(k+2), \; q_0 (k+1)(k+2))
    \]

    so, $p_{\text{foo}(k+1)}$ and $q_{\text{foo}(k+1)}$, denoted as regular $p$ and $q$ would have a ratio of

    \[
    \frac{p}{q}
    =
    \frac{q_0 + p_0 (k+1)(k+2)}
         {q_0 (k+1)(k+2)}
    \text{ where }
    (k+1)p_0 = k q_0
    \text{ or }
    q_0 = \frac{p_0 (k+1)}{k}.
    \]    

    substitude $q_0$ with $\frac{p_0(k+1)}{k}$ as

    \begin{align*}
    \frac{p}{q}
    &= \frac{q_0+p_0(k+1)(k+2)}{q_0 (k+1)(k+2)} \\
    &= \frac{\frac{p_0(k+1)}{k}+p_0(k+1)(k+2)}{\frac{p_0(k+1)}{k}(k+1)(k+2)} \\
    &= \frac{p_0(k+1)\left(\frac{1}{k}+ (k+2)\right)}{p_0\frac{k+1}{k}(k+1)(k+2)} \\
    &= \frac{\frac{1}{k}+ (k+2)}{\frac{k+1}{k}(k+2)} \\
    &= \frac{\frac{1+k(k+2)}{k}}{\frac{k+1}{k}(k+2)} \\
    &= \frac{(k+1)^2}{(k+1)(k+2)} \\
    &= \frac{k+1}{k+2} = 1-\frac{1}{k+2}
    \end{align*}

    since P(k+1) follows P(k), the \textbf{induction case} is satisfied, thus we can conclude the given predicate is true

\end{proof}

\vspace{1em}
\hrulefill

\section*{Task 7: Midway Tower of Hanoi (4 points)}
For this task, save your work in \texttt{smallman/Midway.java}. 

Monks at a remote monastery are busy solving the Tower of Hanoi problem, a duty passed on for tens of generations. According to an old tale, when this group of monks finishes, $P$ will be shown to equal $NP$, and the world will come to an end.

In Tower of Hanoi, you are given $N$ disks labeled $0, 1, \ldots, N-1$ by their sizes. There are 3 pegs: Peg 0, Peg 1, and Peg 2. Initially, all disks are on Peg 0, neatly arranged from small (Disk 0) to large (Disk $N-1$), with the smallest on top. The goal is to transfer all disks to Peg 1. You can move exactly one disk at a time, and a larger disk can never be placed on top of a smaller disk.

Solving this task is straightforward with the following Python program, which prints the fewest number of moves---that is, the optimal solution:
\begin{verbatim}
def solve_hanoi(n, from_peg, to_peg, aux_peg):
    if n>0:
        solve_hanoi(n-1, from_peg, aux_peg, to_peg)
        print("Move disk", n-1, "from Peg", from_peg, "to Peg", to_peg)
        solve_hanoi(n-1, aux_peg, to_peg, from_peg)

solve_hanoi(n, 0, 1, 2)
\end{verbatim}
Following these steps uses $2^N - 1$ moves. Because this number is huge, the monks devised a puzzle: assume they strictly followed the program's instructions. Given an intermediate configuration, determine how many more steps they will need before completing the task. (The original handout includes a visual trace of all configurations generated by the program.)

\textbf{Subtask I:} Prove, using mathematical induction, that for any $n \ge 0$, \texttt{solve\_hanoi(n, ...)} generates exactly $2^n - 1$ lines of instructions.

\medskip
\solution

\begin{proof}

    we claim that

    \[
        P(n) \equiv \text{solve\_hanoi(n, ...) will print out for exactly } 2^n - 1 \text{ lines when } n \ge 0
    \]

    for the \textbf{base case} where $n_0 = 0$

    \[
        P(0) \equiv 2^0 - 1 = 0 \text{ as the function only works when } n > 0 \text{ so nothing got printed out}
    \]

    then, we assume P(k) for arbitrary $(k \ge n_0)$, we prove $P(k+1) = 2^{k+1} - 1$

    \vspace{1em}

    from the recurrence, if $n = 0$, the function does nothing, so nothing got printed out

    \[T(0) = 0\]

    and for $n > 0$, solve\_hanoi(n-1, ...) will print out $T(n-1)$ lines (moves the top n-1 disks to the left or right, so we can move disk n to the middle), then the mother function prints out (move disk n to the middle), and ended with another solve\_hanoi(n-1, ...) (with different parameter), this $T(n-1)$ lines of actually solving for $T(n-1)$ --- which the pile's position will iterate between left and right; thus

    \[T(n) = 2T(n-1) + 1)\]

    we use this recurrence relation to prove $P(k+1)$

    \begin{align*}
        P(k+1) &\equiv T(k+1) = 2T(k) + 1 && \quad \text{from the recurrence}\\
        &= 2(2^k-1) + 1 && \quad \text{induction hypothesis} \\
        &= 2^{k+1} - 2 + 1 = 2^{k+1} - 1
    \end{align*}

    since P(k+1) follows P(k), the \textbf{induction case} is satisfied, thus we can conclude the given predicate is true

\end{proof}
\end{document}
