\documentclass{article}

\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{tikz}
\usepackage[plain]{algorithm}
\usepackage{algpseudocode}

\usetikzlibrary{automata,positioning}

%
% Basic Document Settings
%

\topmargin=-0.45in
\evensidemargin=0in
\oddsidemargin=0in
\textwidth=6.5in
\textheight=9.0in
\headsep=0.25in

\linespread{1.1}

\pagestyle{fancy}
\lhead{\hmwkAuthorName}
\chead{} % empty
\rhead{\hmwkHeaderTitle}
\cfoot{\thepage}

\renewcommand\headrulewidth{0.4pt}
\renewcommand\footrulewidth{0pt}

\setlength\parindent{0pt}

%
% Homework Details
%

\newcommand{\hmwkTitle}{Homework\ 3}
\newcommand{\hmwkDueDate}{Friday, February 6, 2026,\ 11{:}59 pm (Thai time)}
\newcommand{\hmwkClass}{Data Structure\ and Abstractions}
\newcommand{\hmwkClassTime}{Section 2}
\newcommand{\hmwkClassInstructor}{Kanat Tangwongsan}
\newcommand{\hmwkAuthorName}{\textbf{Jiraroj Wiruchpongsanon (6781617)}}

% What appears in the top-right header
\newcommand{\hmwkHeaderTitle}{Data Structure and Abstractions: Homework 3}

%
% Title Page
%

\title{
    \vspace{2in}
    \textmd{\textbf{\hmwkClass:\ \hmwkTitle}}\\
    \normalsize\vspace{0.1in}\small{Due\ on\ \hmwkDueDate}\\
    \vspace{0.1in}\large{\textit{\hmwkClassInstructor\ \hmwkClassTime}}
    \vspace{3in}
}

\author{\hmwkAuthorName}
\date{}

\renewcommand{\part}[1]{\textbf{\large Part \Alph{partCounter}}\stepcounter{partCounter}\\}

%
% Helper Commands
%

\newcounter{partCounter}
\newcommand{\solution}{\textbf{\large Solution}}

\newtheorem*{theorem}{Theorem}

\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\begin{document}

\maketitle
\newpage

\section*{Task 3: Loops \& Numerical Computation (7 points)}

For this task, save your work in hw3.pdf
Consider the Java code below. We have already fleshed out a loop invariant for you. Your task is to
add Hoare logic justifications so that we have a proof that mult returns the product of x and y. You may
be inspired by worked examples from class.

\begin{verbatim}
// precondition: x >= 0 && y >= 0;
    int mult(int x, int y) {
    int k = x, n = y, res = 0;
    while (k != 0) { // @loop_invariant x * y == k * n + res;
        if (k%2 == 1) res = res + n;
        k /= 2;
        n *= 2;
    }
    return res;
}
// post-condition: returns x * y
\end{verbatim}

\medskip
\solution

\vspace{1em}

firstly, precondition are \(x \ge 0\) and \(y \ge 0\). then, on the \textbf{initialization phase}, we got 
\[\(k = x\), \quad \(n = y\), \quad \(res = 0\)\]

, which we could use to define the loop invariant (I) as

\[
I \equiv (x \cdot y = k \cdot n + res) \land k \ge 0 \land n \ge 0
\]
and since
\[
k = x,\ n = y,\ res = 0 \implies kn + res = xy
\]

the loop invariant holds. next, on the \textbf{maintenance phase}, we'll prove 
\[\(\{I \land k \ne 0\}\)\text{ body }\(\{I\}\)\]

\begin{enumerate}
    \item case 1: in case of \(k \in \mathbb{O}\), since \(k\) is odd

    \[
    k = 2v + 1
    \]

    then, after the loop execution, we'll get

    \[
    k' = v, \quad \ n' = 2n, \quad res' = res + n
    \]

    which when combined
    \[
    \begin{aligned}
    k'n' + res' &= v(2n) + (res + n) \\
                &= (2v + 1)n + res && \text{and since } k = 2v + 1\\
                &= kn + res = xy
    \end{aligned}
    \]

    \item case 2: in case of \(k \in \mathbb{E}\)

    then the execution gives
    \[
    k' = v,\ n' = 2n \text{ and } res' = res
    \]

    which combines to

    \[
    \begin{aligned}
    k'n' + res' &= v(2n) + res \\
                &= 2vn + res && \text{since } k = 2v \text{ for even numbers}\\
                &= kn + res = xy
    \end{aligned}
    \]

\end{enumerate}

since both case holds true, the maintenance phase is satisfied, and we reach our \textbf{termination phase}: since the loop invariant holds true before the loop starts, and it still holds true (on both case) when the loops end --- while \(k = 0\).

\vspace{1em}

from

\[
\begin{aligned}
    x \cdot y &= kn + res && \text{ when } k = 0 \\
    &= 0(n) + res
\end{aligned}
\]

thus
\[
x \cdot y = res
\]

the function returns variable `res`, which is equivalent of returning \(x \cdot y\)

\vspace{1em}
\hrulefill

\section*{Task 4: Looping an Array (7 points)}

For this task, save your work in hw3.pdf
The Java code below was written to find the last index that x appears in the array A. If x doesn’t appear in A, then the function returns $−-1$.

\begin{verbatim}
int find(int[] A, int x) {
    int n = A.length, i = n - 1;
    while (i >= 0) {
        if (A[i] == x)
            return i;
        i -= 1;
    }
    return -1;
}
\end{verbatim}
Your task is as follows:
\begin{enumerate}
    \item Write a precondition and a post-condition for find. Your conditions should be in sync with the
invariants and other justification steps you will also make about the code. At the very least, the
post-condition should allow us to say that find does find the last index of x as specified above.
    \item Specify a loop invariant for the loop that find uses.
    \item Fill in Hoare logic justifications so that we have a complete proof that given the precondition, the
post-condition holds, thereby find meets the specification.
\end{enumerate}

\medskip
\solution

\begin{enumerate}
    \item precondition are an array of integer \(A\) that isn't empty (or equals to null --- points to elsewhere but the first member of the array), and \(n\text{ is length of array } A\), \(i\) will be all the indices of the array, ranging from \(0 \text{ to } (n-1)\).

    \vspace{1em}

    post-condition got two cases

    \begin{itemize}
        \item case where the returned value are \(-1\), this means that
        \[
        \forall j \ne x \text{ when } 0 \le j \le n - 1
        \]

        \item or case where the returned value are within the range of \([0, n-1]\), or when the returned value isn't \(-1\). this means that there exist an index \(i\) in array \(A\) that equals to \(x\), and since we search backwards, \(i\) is the last index that \(x\) occours.
    \end{itemize}

    \item \[
    I \equiv \left(\forall j \left(A[j] \ne x \land i < j \le n - 1\right)\right) \land (-1 \le i \le n - 1)
    \]

    \item     
    
    \begin{itemize}
        \item \textbf{Initialization: }
        \[
        n = \text{length of array } A,\ i = n - 1
        \]
        consider the proposition
        \[
        \forall j (i < j < n \Rightarrow A[j] \ne x)
        \]
        but since \(i\) is already equal to \(n - 1\), which gives
        \[
        n - 1 < j < n
        \]
        there is no such \(j\) that could satisfy this inequality, thus there is no counter-example that could be made against this invariant, therefore the invariant holds true.

        \item \textbf{Maintenance: }
        we try to prove
        \[
        \{I \land i \ge 0\}\ \text{body}\ \{I\}
        \]
        \begin{enumerate}
            \item case 1: when \(A[i] = x\).  

            from the invariant we already know that for all indices \(j\) with \(i < j < n\), \(A[j] \ne x\),
            since the current index \(i\) contains \(x\) and no index to the right of \(i\) contains \(x\), index \(i\) is the last occurrence of \(x\), 
            and therefore returning \(i\) satisfies the postcondition.

            \item case 2: when \(A[i] \ne x\).  
            the code executes \(i = i - 1\). Let \(i' = i - 1\)
            for any \(j\) with \(i' < j < n\), either
            \begin{itemize}
                \item \(j = i\), in which case \(A[j] = A[i] \ne x\), or
                \item \(j > i\), in which case \(A[j] \ne x\) by the invariant.
            \end{itemize}
            thus
            \[
            \forall j (i' < j < n \Rightarrow A[j] \ne x)
            \]
            and together with \(-1 \le i' \le n-1\), the loop invariant still holds.
        \end{enumerate}

        \item \textbf{Termination: }  
        the loop ends when \(i < 0\), and from the invariant we also have \(-1 \le i\); therefore at exit \(i = -1\).  
        substituting \(i = -1\) into the invariant
        \[
        \forall j (i < j < n \Rightarrow A[j] \ne x)
        \]
        gives
        \[
        \forall j (0 \le j < n \Rightarrow A[j] \ne x)
        \]
        and we can conclude that \(x\) does not appear anywhere in array \(A\). therefore returning \(-1\) satisfies the postcondition.
    \end{itemize}
\end{enumerate}

\vspace{1em}
\hrulefill

\end{document}
